---
title: "Case Study1 Project"
author: "Lauren Darr, Emmanuel Farrugia, John Heinen, Johnson Ekedum"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: readable
    highlight: textmate
---

##Introduction

The data project seeks to explore world economic and
social indicators of major countries. In this project, our goal is to carry out an exploratory data analysis using two data set, GDP data and Educational data.This data will be downloaded from the world bank website. Original data set are in the links below 

* [GDP data](http://data.worldbank.org/data-catalog/GDP-ranking-table)

* [Education data](http://data.worldbank.org/data-catalog/ed-stats )

##### In this project we will go through a 3 step process which will involve
* Cleaning dataset
* Analysing dataset
* Visualizing data set

To begin we will load the required package and download the dataset into R.

* Loading required package: Please install first if you dont have them
```{r global_options, message=FALSE, warning=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(data.table)
library(Hmisc)
```
Download data from URL
```{r }
GDPurl <-  "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv"
download.file(GDPurl, "GDP.csv", quiet=TRUE)

EduURL <- "https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv"
download.file(EduURL, "Edu.csv", quiet=TRUE)
```

### Part1...Data Cleaning 

Read files into dataframe. Using the "read_csv" from readr package helps eliminate the need to use stringasFactors= F

```{r message=FALSE, warning=FALSE}
GDP.data <- read_csv("GDP.csv",skip =3)  
Educ.data <- read_csv("Edu.csv") 
```
Looking at the first three rows, we see that the GDP data contains  some empty and irrelevant column names 
```{r message=FALSE, warning=FALSE}
head(GDP.data,3)
```
Here we renamed and picked the relevant columns for our analysis
```{r message=FALSE, warning=FALSE}
GDP.data <- GDP.data[, c(1, 2, 4, 5)]
GDP.data <- GDP.data[-c(1:2),]
colnames(GDP.data) <- c("Country_Code", "Rank","Country_Name","GDP_Value")
head(GDP.data,4)
```
* We also converted the character Rank and GDP_Value column in the GDP_data to Numeric and remove commas to aid data manipulation 
```{r message=FALSE,warning=FALSE}
GDP.data$GDP_Value <- as.numeric(str_replace_all(GDP.data$GDP_Value,",",""))
GDP.data$Rank <- as.numeric(GDP.data$Rank)
head(GDP.data$GDP_Value,2)
```
We replaced all empty spaces within Educ_data columns  with an underscore and selected the most relevant for our analysis
```{r message=FALSE, warning=FALSE }
colnames(Educ.data) <- str_replace_all(colnames(Educ.data)," ","_")
Educ.data <- Educ.data %>% 
  select(CountryCode,Long_Name,Short_Name, Income_Group,Region,Currency_Unit)
head(Educ.data,2)
```
Check for the number of  NA's within each column of  GDP datasets

The following output gives the column names and the number of NA's in each column of GDP table
```{r}
sapply(GDP.data,function(x) sum(is.na(x)))
```

Check for the number of NA's within each column of  Education datasets

The following output gives the column names and the number of NA's in each column of  Education table
```{r}
sapply(Educ.data,function(x) sum(is.na(x)))
```
##Part 2 ...Data Analysis
In this data analysis section, we would answer the following questions

Case Study Question 1.**Merge the data based on the country shortcode. How many of the IDs match?**

We merged the two datasets, removed NA's and counted matching ID

As you can see, there are 188 matching ID's after removing NA values.  
```{r}
merge_GDP_EDU <- 
  merge.data.frame(GDP.data,Educ.data, by.x = "Country_Code", by.y = "CountryCode") %>% 
  na.omit()
length(merge_GDP_EDU$Country_Code)
```

Case Study Question 2.**Sort the data frame in ascending order by GDP (so United States is last). What is the 13th country in the resulting data frame?**

After running this analysis, the 13th country in the resulting data frame is St. Kitts and Nevis
```{r}
merge_GDP_EDU <- merge_GDP_EDU %>% 
  select(everything()) %>% 
  arrange(GDP_Value) 
merge_GDP_EDU[13,3]
```

Case Study Question 3.**What are the average GDP rankings for the "High income: OECD" and "High income: nonOECD" groups?** 

Given the output, we see that the average GDP rankings for the High income: nonOECD countries is 91.91304 and 34.06897  for High income: OECD countries respectively
```{r}
Avg_GDP <- merge_GDP_EDU %>% 
  select(everything()) %>% 
  group_by(Income_Group) %>%
  summarise(Average_GDP = mean(Rank))%>%
  filter(Income_Group %in%  c("High income: OECD","High income: nonOECD"))
Avg_GDP
```

Case Study Question 5.**Cut the GDP ranking into 5 separate quantile groups. Make a table versus Income.Group.**
```{r}
merge_GDP_EDU$Quantiles <- cut2(merge_GDP_EDU$Rank, g=5)
table(merge_GDP_EDU$Quantiles, merge_GDP_EDU$Income_Group)

```
+ **How many countries are Lower middle income but among the 38 nations with highest GDP?**
Given the output below, we see that there are 6 countries in the lower middle income but among the 38 nations with the highest GDP.

```{r}
merge_GDP_EDU[151:189,c(1,2,3,4,7)] %>% 
  filter(Income_Group=="Lower middle income")
```
## Part 3...Data Visualization
Case Study Question 4.**Plot the GDP for all of the countries. Use ggplot2 to color your plot by Income Group.**

In this section, we used some discretion in plotting our graph, there are about 189 countries in our dataset. It will not be efficient to plot all countries on the X axis therefore we group the countries by income group and plot against GDP.     
```{r }
ggplot(merge_GDP_EDU, aes(y = GDP_Value, x=Income_Group,fill=Income_Group)) + 
  scale_y_log10()+ 
  geom_point(pch = 21, size = 8, stat = "identity", position=position_jitter())+
  scale_fill_manual(values = c("red", "orange", "green", "blue","brown"),
                    na.value = "grey50" ) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Conclusion.
According to  The Organization for Economic Cooperation and Development.
"OECD member countries account for 63 percent of world GDP, three-quarters of world trade, 95 percent of world official development assistance." 


During our analysis, it was interesting to discover that there are countries that do not fall within the OECD and high income bracket  whose GDP is greater than majority of high income  and high income OECD countries. To name a few , we see countiries such as China, India, Indonesia, Nigeria and some others whose GDP is growing by leaps and bounds. 

There is  no doubt that these middle income countries given the consistent rise in their GDP will one day become the engine of economic growth, prosperity, and sustainable development.


